---
title: "Storm Data Events Analytics: Analysis of storms and other significant weather phenomena
        across US"
author: "Rama Tripathy"
date: "March 12, 2017"
output: html_document
---

## Synopsis  

In this report, we analyze the Storms data collected from National Weather Services(NWS) and other sources including media reports and try to answer these questions:  
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?  

## Setup options
### Software Environment  
1. R version 3.0.2  
2. Platform: x86_64-windows  
3. Attached base packages:  
- data.table package to directly read zip/.bz2/csv format      
- plyr and dplyr to manipulate the data    
- ggplot2 package for plots    
- xtable package to produce nice tables    
-   
4. Setup local directory
5. Setup global options  
```{r}
knitr::opts_knit$set(warn=-1) #surpress warnings
knitr::opts_knit$set(echo = TRUE)
library(knitr)
library("data.table")
library(plyr)
library(dplyr)
library(ggplot2)
library(xtable)
# install package 'fiftystater' from github
# install.packages("devtools")
# devtools::install_github("wmurphyrd/fiftystater")
library(fiftystater)

knitr::opts_knit$set(root.dir = "E:\\Coursera\\Reproducible_Research\\Storms_Data_Analytics\\")
```

## Data Processing  
###  Read raw Data from web site  
```{r}
knitr::opts_knit$set(warn=-1) #surpress warnings
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url,destfile="repdata-data-StormData.csv.bz2")
stormdata <- read.table("StormData.csv.bz2", header=TRUE, quote="\"", sep=",", 
                        stringsAsFactors = FALSE)
```
### Check dimension and the variables and their types   
```{r}
glimpse(stormdata)
```
### Take an intial look into the storm data  
```{r}
head(stormdata)
```
### Check summary to:  
- identify  missing values  
- minimum and maximum field values to determine possible outliers  
```{r}
summary(stormdata)
```
- Field F (Turnado F scale) has NA's :843563   
- Field LATITUDE has NA's   :47    
- Field LONGITUDE has NA's   :40   
- We can use Imputation tools like Mice, HIMSC to impute missing values    
- But for our analysis, we will not be using these fields, so ignoring them  

### Identify fields to be used for our analysis  
1. We will need the following fields:  
Location - STATE, COUNTY, LATITUDE, LONGITUDE, LATITUDE_E, LONGITUDE_)  
Event type (EVTYPE)  
Beginning/end date and time (BGN_DATE, BGN_TIME, TIME_ZONE, END_DATE, END_TIME)  
Health Impact - FATALITIES, INJURIES  
Property Damage - comprised of two parts:  
- a 3-significant numeric field (PROPDMG)  
- a field to indicate magnitude (K = 1000, M = Million...) PROPDMGEXP  
Corp Damage - comprised of two parts:  
- a 3-significant numeric field (CROPDMG)  
- a field to indicate magnitude (K = 1000, M = Million...) CROPDMGEXP  
```{r}
stormdata <- select(stormdata, STATE, COUNTY, LATITUDE, LONGITUDE, 
                    LATITUDE_E, LONGITUDE_, EVTYPE, BGN_DATE, BGN_TIME, 
                    TIME_ZONE, END_DATE, END_TIME, FATALITIES, INJURIES,
                    PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REFNUM)
```

2. Transform STATE, EVTYPE to factor variables
```{r}
stormdata$STATE <- as.factor(stormdata$STATE)
stormdata$EVTYPE <- as.factor(stormdata$EVTYPE)
stormdata$PROPDMGEXP <- as.character(stormdata$PROPDMGEXP)
stormdata$CROPDMGEXP <- as.character(stormdata$CROPDMGEXP)
```

### Calculate actual Property/Crop Damages in thousands 
1. We will convert the magnitude field to numeric value and multiply it with  
property/crop damage field to calculate actual vale.  
2. PROPDMGEXP column specifies the amount for PROPDMG in:  
- Hundreds(H/0/1/2/3/4/5/6/7/8)  
- Thousands(K)  
- Millions(M)  
- Billions(B)
- consider blank/-/+/? values as 10 for PROPDMGEXP  

3. CROPDMGEXP column specifies the amount for CROPDMG in:  
- Hundreds(0/1/2)  
- Thousands(K)  
- Millions(M)  
- Billions(B)
- consider blank/-/+/? values as 10 for CROPDMGEXP
4. We will use switch function for this conversion  
```{r}
magnitudeValue <- function(value, magnitude){
             v <- switch(magnitude,
                      H = .1, h = .1,
                      K = 1, k = 1, 
                      M = 1000, m = 1000, 
                      B = 1000000, b = 1000000, 
                      Otherwise = .01)
             if(!is.null(v))
               value <- value*v
             else  value <- .01*value
             value
             }
```
5. Calculate property damage, crop damage and total damage  
total economic impact= property damage + crop damage  
```{r}
knitr::opts_knit$set(warn=-1) #surpress warnings
stormdata <- as.data.frame(stormdata %>% group_by(REFNUM) %>% mutate(PropertyDamage=magnitudeValue(PROPDMG,PROPDMGEXP)))
stormdata <- as.data.frame(stormdata %>% group_by(REFNUM) %>% mutate(CropDamage=magnitudeValue(CROPDMG,CROPDMGEXP)))
stormdata$TotalDamage <- stormdata$CropDamage + stormdata$PropertyDamage
```
### Calculate total health impact
total health impact= fatalities + injuries
```{r}
stormdata$TotalHealthImpact <- stormdata$FATALITIES + stormdata$INJURIES
```
### Take alook into the new fields in  storm data now
```{r}
head(select(stormdata, FATALITIES,INJURIES,TotalHealthImpact,
             PropertyDamage,CropDamage,TotalDamage))
```
Free up some memory  
```{r}
gc(verbose = FALSE)
```
## Results
### Analysis across US
### Which types of events are most harmful to population health?
Aggregate and find most harmful event type to health
```{r}
knitr::opts_knit$set(cache = TRUE)
# Number of injuries by eventtype
injuriesbyev <- aggregate(INJURIES ~ EVTYPE, data = stormdata, FUN = sum)
#
# Number of fatalities by event type
fatalitiesbyev <- aggregate(FATALITIES ~ EVTYPE, data = stormdata, FUN = sum)
#
# Total Health Impact by event type
hibyev <- merge(injuriesbyev, fatalitiesbyev) # merge by event tye
# Calculate the total health loss by eventtype
hibyev <- mutate(hibyev, totalbyev = INJURIES + FATALITIES ) %>% setorder(-totalbyev) 
rm(injuriesbyev)
rm(fatalitiesbyev)
```
1. Most harmful events to population health are:  
```{r}
library(xtable)
kable(head(hibyev,20))
```
2. Plot Top 10 most harmful events data using ggplot2 - using bar charts  
```{r}
library(ggplot2)
ggplot(hibyev[1:10,], aes(x=EVTYPE, y = totalbyev, fill = EVTYPE)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) + 
  xlab("Event Type") + 
  ylab("Casualities") +
  ggtitle("Most Harmful Events to population health")
```
### Which types of events have the greatest economic consequences across US?
Aggregate and find most harmful event type to economy
```{r}
knitr::opts_knit$set(cache = TRUE)
# Property damage by eventtype
pdmgbyev <- aggregate(PROPDMG ~ EVTYPE, data = stormdata, FUN = sum)
# Number of fatalities by event type
cdmgbyev <- aggregate(CROPDMG ~ EVTYPE, data = stormdata, FUN = sum)
tdmgbyev <- merge(pdmgbyev, cdmgbyev) # merge by event tye
# Calculate the total damage by eventtype
tdmgbyev <- mutate(tdmgbyev, TotalDamage = PROPDMG + CROPDMG ) %>% setorder(-TotalDamage)
rm(pdmgbyev)
rm(cdmgbyev)
```
1. Top events having greatest economic consequences:
```{r}
library(xtable)
kable(head(tdmgbyev,20))
```
2. Plot Economic harmful events across US data using ggplot2 - using bar charts  
```{r}
library(ggplot2)
ggplot(tdmgbyev[1:10,], aes(x=EVTYPE, y = TotalDamage, fill = EVTYPE)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) + 
  ylab("Total Damage in thousand") + 
  xlab("Event Type") +
  ggtitle("Most Economic Harmful Events across US")
```
### Analysis of Storm Even Impact by each State in US  
1. Aggregate and find most harmful to health impact by state  
```{r}
knitr::opts_knit$set(cache = TRUE)
# Number of injuries by eventtype by stste
injuriesbyst <- aggregate(INJURIES ~ EVTYPE + STATE, data = stormdata, 
                          FUN = sum)
#
# Number of fatalities by event type
fatalitiesbyst <- aggregate(FATALITIES ~ EVTYPE + STATE, data = stormdata, 
                            FUN = sum)
#
# Total Health Impact by event type
hibyevst <- merge(injuriesbyst, fatalitiesbyst) #merge by event type and state
# Calculate the total health loss by eventtype
hibyevst <- mutate(hibyevst, totalbyst = INJURIES + FATALITIES ) %>% setorder(-totalbyst) 
rm(injuriesbyst)
rm(fatalitiesbyst)
```
2. Aggregate and find most harmful event type to economy by state 
```{r}
knitr::opts_knit$set(cache = TRUE)
# Property damage by eventtype
pdmgbyst <- aggregate(PROPDMG ~ EVTYPE + STATE, data = stormdata, FUN = sum)
# Number of fatalities by event type
cdmgbyst <- aggregate(CROPDMG ~ EVTYPE + STATE, data = stormdata, FUN = sum)
tdmgbyevst <- merge(pdmgbyst, cdmgbyst) # merge by event type and state
# Calculate the total damage by eventtype
tdmgbyevst <- mutate(tdmgbyevst, TotalDamage = PROPDMG + CROPDMG ) %>% setorder(-TotalDamage)
rm(pdmgbyst)
rm(cdmgbyst)
```



